---
title: "CUNY - DS Maths - Final Project"
author: "Kishore Prasad"
date: "November 27, 2015"
output: html_document
---

Lets first prepare the environment to use hflights data

```{r}
#install.packages("hflights")
library(hflights)
```

Assuming that my X variable is DepDelay and Y variable is ArrDelay,  we calculate the 3rd Quartile of DepDelay and 2nd Quartile of ArrDelay as below

```{r}
DD_3Q <- quantile(hflights$DepDelay, na.rm = TRUE)[4]
AD_2Q <- quantile(hflights$ArrDelay, na.rm = TRUE)[3]

```

We now, create 2 additional columns that count the observations for being greater than or less than the 3Q and 2Q variables.

```{r}

fdata <- cbind(hflights[ ,c("DepDelay", "ArrDelay")], A = ifelse(hflights$DepDelay>DD_3Q,1,0), B = ifelse(hflights$ArrDelay>AD_2Q,1,0) )

tbl <- table(fdata$A, fdata$B)
tbl
```


Using this table, we now generate the probabilities table as below:

```{r}
p<- tbl[1,1]
q<- tbl[1,2]

r<- tbl[2,1]
s<- tbl[2,2]

prob_table <- matrix(data=0,nrow=3, ncol=3)

prob_table[1,1] <- p / (p+q+r+s) 
prob_table[1,2] <- q / (p+q+r+s) 
prob_table[1,3] <- (p+q) / (p+q+r+s) 

prob_table[2,1] <- r / (p+q+r+s) 
prob_table[2,2] <- s / (p+q+r+s) 
prob_table[2,3] <- (r+s) / (p+q+r+s) 

prob_table[3,1] <- (p+r) / (p+q+r+s) 
prob_table[3,2] <- (q+s) / (p+q+r+s) 
prob_table[3,3] <- (p+q+r+s) / (p+q+r+s) 

prob_table

```

Using the prob_table, we will now answer the following questions:

1. P(X>x | Y>y)

```{r}

prob_table[2,2]

```


2. P(X>x and Y>y)

```{r}

prob_table[2,3] * prob_table[3,2] 

```

3. P(X<x | Y>y)

```{r}

prob_table[1,2] 

```

4. P(X<x and Y>y)

```{r}

prob_table[1,3] * prob_table[3,2] 

```

generate the chisquare test

```{r}
chisq.test(tbl)
```

We can see from the above chi square output that the p-value is less than 0.05. we reject the null hypothesis that the Departure delay is independent of the arrival delay. 


2. Descriptive and Inferential Statistics.

```{r}
library(psych)
describe(fdata)

hist(fdata$DepDelay)
hist(fdata$ArrDelay)

```

A scatterplot of the 2 variables is as below

```{r}

plot(fdata$DepDelay, fdata$ArrDelay)

```

We will now generate the correlation matrix

```{r}

mat1 <- matrix(table(fdata$A), nrow=1, ncol=2)
mat2 <- matrix(table(fdata$B), nrow=1, ncol=2)
cor_mat <- cor(mat1,mat2) 
cor_mat

```


Provide a 95% CI for the difference in the mean of DepDelay and ArrDelay.

```{r}

N=nrow(fdata)
mean_depDelay <- mean(fdata$DepDelay, na.rm = TRUE)
mean_arrDelay <- mean(fdata$ArrDelay, na.rm = TRUE)

var_depDelay <- var(fdata$DepDelay, na.rm = TRUE)
var_arrDelay <- var(fdata$ArrDelay, na.rm = TRUE)

MSE <- (var_depDelay + var_arrDelay) / 2

sm1_m2 <- sqrt(2 * MSE / N)
sm1_m2
mean_diff <- mean_depDelay - mean_arrDelay
mean_diff
tCL <- 2.074

lower_limit_95 <- mean_diff - (tCL * sm1_m2)
lower_limit_95
upper_limit_95 <- mean_diff + (tCL * sm1_m2)
upper_limit_95

```

We will now generate the correlation matrix for the variables depdelay and arrdelay as follows:

```{r}

cor_mat <- cor(fdata[1:2], use="complete.obs", method = "pearson")
cor_mat

library(corrplot)
corrplot(cor_mat, type="upper", order="hclust", tl.col="black", tl.srt=45)

```

As we can see from the above correlation matrix, there is a strong positive correlation between depdelay and arrdelay. 

```{r}

cor_mat
prec_mat <- solve(cor_mat)
prec_mat

A <- cor_mat
B <- prec_mat

AB <- A %*% B
AB

BA <- B %*% A
BA

```

Shift values so that the minumum value is above zero

```{r}

fdata <- cbind(fdata, depDelay_shift = fdata$DepDelay - min(fdata$DepDelay, na.rm=TRUE), arrDelay_shift = fdata$ArrDelay - min(fdata$ArrDelay, na.rm=TRUE) )

fdata$DepDelay <- fdata$DepDelay - min(fdata$DepDelay, na.rm=TRUE)

```

Lets now load the MASS package and fit an exponential pdf.

```{r}

library(MASS)

depdelay_fit <- fitdistr(fdata[is.na(fdata$depDelay_shift)==FALSE,"depDelay_shift"], "exponential")
```


Next find the optimal value of lamdba and take 1000 samples

```{r}
depdelay_fit <- fitdistr(fdata[is.na(fdata$depDelay_shift)==FALSE,"depDelay_shift"], "poisson")

depdelay_fit
lambda <- as.numeric(depdelay_fit[1])

depdelay_samples <- rexp(1000, lambda)

```

Next we compare the histogram of the original variable with the sample

```{r}

hist(fdata$depDelay_shift)
hist(depdelay_samples)

```